Microlensing as a possible probe of event-horizon structure in quasars

### 2016-10-20 paper1/ms.tex
corrections

\begin{equation} \mu_{\rm total} = \sum_{i} \mu(\vec\theta_i)
\end{equation}
applies. It is possible for the magnification to become formally
infinite, as a result of an eigenvalues of the magnification matrix
(\ref{eqn:magnif-matrix}) becoming infinite.  This generically happens
on curves on the $\vec\theta$ plane, known as {\em critical curves}.
Mapping a critical curve to the source plane, through the lens
equation (\ref{eqn:lens}), to the source plane $\vec\beta$ gives the
so-called caustics.  Caustics can appear in the optical system, not just
gravitational lensing.  For a point source, caustics are singularities
of the magnification; for finite-size sources caustics correspond to
high and sharply changing magnification.

@@

In the present study we are analysing three types of sources with different surface brightness: 
\begin{enumerate}
 \renewcommand{\theenumi}{(\arabic{enumi})}
  \item a rotationally symmetric source with a bivariate gaussian surface brightness distribution,
  \item a disk source with constant surface brightness distribution,
  \item a crescent shaped source with constant surface brightness distribution.
\end{enumerate}
The first two sources are the typical choices used in the literature to describe the luminous parts of a quasar (Prasenjit should give some citations here). The third one is a recently proposed
variant \citep{2013MNRAS.434..765K}.

@@

In figure \ref{fig:char_points} There are four characteristic points visible on the resulting one-dimensional light profile of the crescent shaped source. Two of the points, p1 and p4, mark the outer boundaries of the luminous disc component. The other two, p2 and p3, mark the boundaries of the dark disc component. Since $S_{1D}$ is a projection of $S_{2D}$ on a line perpendicular to the caustic, the following relation holds:
\begin{equation}
    p_2-p_1 = R_p -R_n - a.
\end{equation}
In addition there are two other obvious relations the characteristic points which are independent of the projection:
\begin{equation}
    p_4 -p_1 = 2 R_p,
\end{equation}

@@

Due to the nature of the caustic and the monotonic behaviour of the magnification map on both sides of the caustic. The previously mentioned characteristic points are inherited by the microlensing lightcurve. The points on the temporal dimension $t_1, t_2, t_3$ and $t_4$ correspond to instances in time when the fold is aligned with $p_1, p_2, p_3$ and $p_4$, respectively. For a constant relative velocity $v_p$ between the source and the caustic there is a simple relation between the points $p_i$ and instances $t_i$:

@@

Figure \ref{fig:lightcurve_gauss} reveals that the 
lightcurve of a crescent source has more
visible features than the other two light-curves corresponding to the
disc and gaussian shape. There are three regions where $S_{1D}$ and by
inheritance, the lightcurve has distinguishable behaviour. The first
region that would be recorded on a lightcurve plot represents the
period of time when the bright disk begins to be overlapped by the
caustic and stops when just before the dark disk reaches the
caustic. During this period of time, the flux of light from the source
is increasingly magnified. The second period starts and ends with the
overlapping of the dark disc. As a boundary of the two regions, there
is a distinguishable point where the slope of the magnification is
drastically changed. This apparent discontinuity in the first
derivative of the magnification function is caused by the caustic
amplification of the sudden drop in the $S_{1d}$ function. During the
respective period, the magnification growth slows or even reverses
during the first part of the period and starts to grow faster again as
the dark disk ends its overlap with the caustic. At the point where
the dark disc clears the caustic, the growth of the magnification is
infinite, which appears as a saddle point on the lightcurve. Next, the
final period corresponds to the case when the dark disc has cleared
the caustic and the bright disc continues to overlap with the
caustic. During this period, the lightcurve reaches a peak that for
most of the parameter space is global and for the rest of the
parameter space local.  At the end of the period, the growth of the
magnification is negative infinite. the respective point appears as a
second saddle point on the lightcurve. Past this point the
magnification of any finite source will decay in roughly the same
manner.

The impact of parameter $a$ on the shape of the lightcurve can be
observed in Figure~\ref{fig:a_var}. Sources where the center of the
dark disc reaches the caustic before the center of the bright disc are
characterized by a smoother broad peak in contrast to the cases where
the center of the dark disc reaches the caustic after the bright disk.
In the case of the latter the instance when the dark disc reaches the
caustic corresponds to larger and larger magnifications until it
becomes a local and even global peak. The effect of the $R_n$
parameter on the light curve is presented in
Figures~\ref{fig:lightcurve_crescent} and
\ref{fig:lightcurve_crescent_back}. For the particular set of
parameters where $R_p = R_n +a$ the third period of time discussed
previously does not exist. Particularizing further, if the value of
the radius of the dark disc is comparable to the value of the radius
of the bright one then the position and shape of the maximum
magnification are strikingly different. In case the crescent reaches
the caustic with the bright region first, the peak magnification
happens when the dark region reaches the caustic and it is
characterized by a sharp variation in magnification growth, In the
opposite case, the peak appears before the end of the bright disc
reaches the caustic and shape is smoother.

@@

For the purpose of this analysis, it was desirable to mimic the
analytical behaviour of a simple fold as much as possible, for
comparing the numerical result with the analytical one, therefore, the
path of the source was chosen, so that it intersects the border of the
caustic perpendicularly, and on a point where the border is a fold
caustic.  We choose the line AC in Figure~\ref{fig:magnification_map}
and consider its two caustic-crossing segments separately:
\begin{itemize}
\item ``path~I'' runs from C to B;
\item ``path II'' runs from A to B.
\end{itemize}
The magnification along paths I and II are very similar, but not
identical.

@@

Note that fitting involves three nuisance parameters, namely the
brightness normalisation and the start- and endpoints on the path.
The latter two are equivalent to considering the location of the
caustic with respect to the starting point and the speed of the
source.  The fitting method automatically marginalises over all the
nuisance parameters.

@@

Figure \ref{fig:gc_forward} shows the result of attempting to a
Gaussian source.  There is a small but statistically very significant
misfit because a Gaussian curve cannot reproduce the part of the
lightcurve with the slight flattening and then a second sub-peak
(cf.~Figure~\ref{fig:char_points}) as the dark part of the crescent
crosses the caustic.

@@

The first two source profiles that we consider are the uniform disk and symmetric Gaussian source. 
Both of them can be described by a half-light radius ($r_{1/2}$) and a total unlensed light flux ($S_0$). 
With the two parameters constrained the one-dimensional profiles, as well as the lightcurves of the 
two source, are completely determined, since no free parameter remains.

@@

Nevertheless, the respective displacement is relevant for the calculation of
of the half-light radius.  It's qualitative effect is to 
decrease the value of the
half-light radius when the absolute value of the parameter is
increased.  In contrast to the simpler source shapes discussed before the
crescent has two additional relevant parameters: the radius of the dark 
disk ($R_n$) and the center displacement in the direction
 perpendicular to the caustic $a$.
Figure 4 reveals that the lightcurve of a crescent source has more
visible features than the other two light-curves corresponding to the
disc and Gaussian shape. The parameters $R_n$ and $a$ have strong
influences on the shape of the microlensed lightcurve as can be seen
in Figures~\ref{fig:a_var}, \ref{fig:lightcurve_crescent} and
\ref{fig:lightcurve_crescent_back}.  Moreover, the one-dimensional
source profile corresponding to the direction perpendicular to the
caustic reveal four characteristic points. The overlap of each of
these points with the caustic leaves visible features on the
lightcurve at the corresponding instances of time. In timely order,
the instances correspond to the start of the overlap between the
caustic and the bright disk, the start of the overlap between the
caustic and the dark disk, the end of the overlap between the caustic
and dark disk and finally the end of the overlap between the caustic
and the bright disk.

>>>>

\begin{equation} \mu_{\rm total} = \sum_{i} \mu(\vec\theta_i)
\end{equation}
applies. It is possible for the magnification to become formally
infinite, as a result of an eigenvalue of the magnification matrix
(\ref{eqn:magnif-matrix}) becoming infinite.  This generically happens
on curves on the $\vec\theta$ plane, known as {\em critical curves}.
Mapping a critical curve to the source plane, through the lens
equation (\ref{eqn:lens}), to the source plane $\vec\beta$ gives the
so-called caustics.  Caustics can appear in the optical system, not just
during gravitational lensing.  For a point source, caustics are singularities
of the magnification; for finite-size sources caustics correspond to
high and sharply changing magnification.

@@

In the present study we are analysing three types of sources with different surface brightness: 
\begin{enumerate}
 \renewcommand{\theenumi}{(\arabic{enumi})}
  \item a rotationally symmetric source with a bivariate gaussian surface brightness distribution,
  \item a disk source with constant surface brightness distribution,
  \item a crescent shaped source with constant surface brightness distribution.
\end{enumerate}
The first two sources are the typical choices used in the literature to describe the luminous parts of a quasar. 
The third one is a recently proposed variant \citep{2013MNRAS.434..765K}.

@@

In figure \ref{fig:char_points} there are four characteristic points visible on the resulting one-dimensional 
light profile of the crescent shaped source. Two of the points, p1 and p4, mark the outer boundaries of the
 luminous disc component. The other two, p2 and p3, mark the boundaries of the dark disc component. 
Since $S_{1D}$ is a projection of $S_{2D}$ on a line perpendicular to the caustic, the following relation holds:
\begin{equation}
    p_2-p_1 = R_p -R_n - a.
\end{equation}
In addition there are two other obvious relations the characteristic points which are independent of the projection:
\begin{equation}
    p_4 -p_1 = 2 R_p,
\end{equation}

@@

Due to the nature of the caustic and the monotonic behaviour of the magnification map on both sides of the caustic, the previously 
mentioned characteristic points are inherited by the microlensing lightcurve. The points on the temporal dimension 
$t_1, t_2, t_3$ and $t_4$ correspond to instances in time when the fold is aligned with $p_1, p_2, p_3$ and $p_4$, respectively. 
For a constant relative velocity $v_p$ between the source and the caustic there is a simple relation between the points $p_i$ and instances $t_i$:

@@

Figure \ref{fig:lightcurve_gauss} reveals that the 
lightcurve of a crescent source has more
visible features than the other two light-curves corresponding to the
disc and gaussian shape. There are three regions where $S_{1D}$ and by
inheritance, the lightcurve has distinguishable behaviour. The first
region that would be recorded on a lightcurve plot represents the
period of time when the bright disk begins to be overlapped by the
caustic and stops when just before the dark disk reaches the
caustic. During this period of time, the flux of light from the source
is increasingly magnified. The second period starts and ends with the
overlapping of the dark disc. As a boundary of the two regions, there
is a distinguishable point where the slope of the magnification is
drastically changed. This apparent discontinuity in the first
derivative of the magnification function is caused by the caustic
amplification of the sudden drop in the $S_{1d}$ function. During the
respective period, the magnification growth slows or even reverses
and later starts to grow faster again as
the dark disc ends its overlap with the caustic. At the point where
the dark disc clears the caustic, the growth of the magnification is
infinite, which appears as a saddle point on the lightcurve. Next, the
final period corresponds to the case when the dark disc has cleared
the caustic and the bright disc continues to overlap with the
caustic. During this period, the lightcurve reaches a peak that for
most of the parameter space is global and for the rest of the
parameter space local.  At the end of the period, the growth of the
magnification is negative infinite. The respective point appears as a
second saddle point on the lightcurve. Past this point the
magnification of any finite source will decay in roughly the same
manner.

The impact of parameter $a$ on the shape of the lightcurve can be
observed in Figure~\ref{fig:a_var}. Sources where the center of the
dark disc reaches the caustic before the center of the bright disc are
characterized by a smoother broad peak in contrast to the cases where
the center of the dark disc reaches the caustic after the bright disk.
In the case of the latter the instance when the dark disc reaches the
caustic corresponds to larger and larger magnifications until it
becomes a local and even global peak. The effect of the $R_n$
parameter on the light curve is presented in
Figures~\ref{fig:lightcurve_crescent} and
\ref{fig:lightcurve_crescent_back}. For the particular set of
parameters where $R_p = R_n +a$ the third period of time discussed
previously does not exist. Particularizing further, if the value of
the radius of the dark disc is comparable to the value of the radius
of the bright one then the position and shape of the maximum
magnification are strikingly different. In case the crescent reaches
the caustic with the bright region first, the peak magnification
happens when the dark region reaches the caustic and it is
characterized by a sharp variation in magnification growth. In the
opposite case, the peak appears before the end of the bright disc
reaches the caustic and shape is smoother.

@@

For the purpose of this analysis, it was desirable to mimic the
analytical behaviour of a simple fold as much as possible. For
comparing the numerical result with the analytical one, therefore, the
path of the source was chosen, so that it intersects the border of the
caustic perpendicularly, and on a point where the border is a fold
caustic.  We choose the line AC in Figure~\ref{fig:magnification_map}
and consider its two caustic-crossing segments separately:
\begin{itemize}
\item ``path~I'' runs from C to B;
\item ``path II'' runs from A to B.
\end{itemize}
The magnification along paths I and II are very similar, but not
identical.

@@

Note that fitting involves three nuisance parameters, namely the
brightness normalisation and the start and end points on the path.
The latter two are equivalent to considering the location of the
caustic with respect to the starting point and the speed of the
source.  The fitting method automatically marginalises over all the
nuisance parameters.

@@

Figure \ref{fig:gc_forward} shows the result of attempting to fit a
Gaussian source.  There is a small but statistically very significant
misfit because a Gaussian curve cannot reproduce the part of the
lightcurve with the slight flattening and then a second sub-peak
(cf.~Figure~\ref{fig:char_points}) as the dark part of the crescent
crosses the caustic.

@@

The first two source profiles that we consider are the uniform disk and the symmetric Gaussian source. 
Both of them can be described by a half-light radius ($r_{1/2}$) and a total unlensed light flux ($S_0$). 
With the two parameters constrained the one-dimensional profiles, as well as the lightcurves of the 
two source, are completely determined, since no free parameter remains.

@@

Nevertheless, the respective displacement is relevant for the calculation of
of the half-light radius.  It's qualitative effect is to 
decrease the value of the
half-light radius when the absolute value of the parameter is
increased.  In contrast to the simpler source shapes discussed before the
crescent has two additional relevant parameters: the radius of the dark 
disk ($R_n$) and the center displacement in the direction
 perpendicular to the caustic $a$.

Figure 4 reveals that the lightcurve of a crescent source has more
visible features than the other two light-curves corresponding to the
disc and Gaussian shape. The parameters $R_n$ and $a$ have strong
influences on the shape of the microlensed lightcurve as can be seen
in Figures~\ref{fig:a_var}, \ref{fig:lightcurve_crescent} and
\ref{fig:lightcurve_crescent_back}.  Moreover, the one-dimensional
source profile corresponding to the direction perpendicular to the
caustic reveal four characteristic points. The overlap of each of
these points with the caustic leaves visible features on the
lightcurve at the corresponding instances of time. In timely order,
the instances correspond to the start of the overlap between the
caustic and the bright disk, the start of the overlap between the
caustic and the dark disk, the end of the overlap between the caustic
and dark disk and finally the end of the overlap between the caustic
and the bright disk.
